---
title: "Bat Montoring Site Analysis"
author: "Jialun Darren Huang"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

### Setup
```{r, message = FALSE}
library(readr)
library(dplyr)
library(knitr)
library(stringr)
library(lubridate)
library(MVN)
library(forecast)
```

The library packages are as follows

* readr: Used to read csv files
* dplyr: Used to manipulate dataset
* knitr: Used to display dataframes in formatted tables
* stringr: Used to manipulate strings
* lubridate: Used to convert and manipulate date and time
* MVN: Used to find multivariate outliers
* forecast: Used to manipulate values using Box-Cox Transformation

### Executive Summary

In this project, we look into how we can use preprocessing techniques to improve the data quality and integrity of a dataset. We will also look into merging dataset techniques as we use two datasets that have similar variables and are related to each other.

The procedure is as follows:

- Reading the Datasets
  - We will be importing the datasets into R
- Subseting Observations
  - Instead of using the full dataset, we will take a sample by subsetting the number of observations
- Merging Datasets
  - Combining datasets with similar variable names that have similar values
- Tidy Dataset
  - After the dataset is merged, we can change the structure from an untidy format (wide) to tidy format (long)
- Variable Creation/Mutation
  - Create new variables from original variables
- Understand the Dataset
  - Once we have merged and tidy the dataset, we can take a closer look at the structure, convert variable data types and relabelling variables where necessary to improve the dataset quality
- Scanning Issues in Dataset
  - Identifying missing, special and inconsistent values to fix data integrity
  - Identifying univariate and multivariate outliers and either impute or remove them
- Data Transformation
  - Applied transformation methods like natural log on a specific variable to reduce skewness.

Finally, we discuss any challenges faced, ease and strengths of what we found easy to do, and what we have learnt from this whole project.

### Data Description
In this project, we will be looking at two datasets that monitor and record temperature and humidity from bat monitoring sites.

Dataset 1 Link: [Temperature and Humidity (Rising HF sensors) CSV](https://discover.data.vic.gov.au/dataset/temperature-and-humidity-rising-hf-sensors/resource/14200bdb-fb16-51ba-a312-afac7f89a3d5)

Dataset 2 Link: [Temperature and Humidity for Bats monitoring CSV](https://discover.data.vic.gov.au/dataset/temperature-and-humidity-for-bats-monitoring/resource/c501ce88-bf15-5e6a-b41a-3e1a145091a3)

### Read/Import Data

<To be written>

```{r, message = FALSE}
rhf_df = read_csv("rising-hf.csv")
bats_df = read_csv("temperature-and-humidity-monitoring-for-bats.csv")
kable(head(rhf_df[, 1:8], n = 3))
kable(head(rhf_df[, 9:length(colnames(rhf_df))], n = 3))
kable(head(bats_df, n = 3))
```

We can briefly visualise the variables and a sample observation from each dataset. We will look more at the variables and structure later.

### Subset Observations

We will subset both datasets to 80% of the data and set the random seed. This ensures that the results from the following code below can be reproduced and consistent.

```{r, message = FALSE}
set.seed(3755729)
```

We create a function `subset_df()`, to subset the dataset. You can control the percentage of subsetted data with `subset_perc`, with the default value at 80%.

```{r, message = FALSE}
subset_df <- function(df, subset_perc = 0.8) {
  rows <- ceiling(subset_perc * nrow(df))
  subset_df <- df[sample(nrow(df), rows), ]
  return(subset_df)
}
```

```{r, message = FALSE}
subset_rhf_df <- subset_df(rhf_df)
cat("Dim: ", dim(subset_rhf_df))

subset_bats_df <- subset_df(bats_df)
cat("Dim: ", dim(subset_bats_df))
```

Even after subsetting the datasets, the dataset is very large. We can subset it further.
Lets further subset it to 1% from the subsetted dataset.

```{r, message = FALSE}
subset_rhf_df <- subset_df(subset_rhf_df, subset_perc = 0.01)
cat("Dim: ", dim(subset_rhf_df))

subset_bats_df <- subset_df(subset_bats_df, subset_perc = 0.01)
cat("Dim: ", dim(subset_bats_df))
```

We now have a subsetted dataset with 7225 observations and 11 variables, and another subsetted dataset with 4356 observations and 8 variables.

### Merge Datasets

Next, we can merge the datasets together, using `bind_rows()`.

```{r, message = FALSE}
subset_merged_df <- bind_rows(subset_rhf_df, subset_bats_df)
```

### Tidy & Manipulate Data I (Tidy Dataset)

Tidy principles
- Each variable must have its own column.
- Each observation must have its own row.
- Each value must have its own cell.

Both datasets are untidy, as the `location` variable contains more than one value. In order for the dataset to be tidy, we have to separate the values to `lat` and `long`

`extract_location_values()` function extracts the `location` values to `lat` and `long`. 
Note that there can be missing values in `location`. For now we will set both `lat` and `long` to also NA. We will deal with the missing values later.

```{r, message = FALSE}
extract_location_values <- function(loc_val) {
  if (is.na(loc_val)) {
    return(c(NA_real_, NA_real_))
  }
  location <- strsplit(loc_val, ",")[[1]]
  lat <- as.numeric(str_trim(location[1]))
  long <- as.numeric(str_trim(location[2]))
  return(c(lat, long))
}
```

`convert_location_lat_long()` function uses the matrix array output from `extract_location_values()` and populates the dataframe with `lat` and `long` variables

```{r, message = FALSE}
convert_location_lat_long <- function(df) {
  lat_long <- sapply(df$location, extract_location_values)
  colnames(lat_long) <- NULL
  df$lat <- lat_long[1, ]
  df$long <- lat_long[2, ]
  return(df)
}
```

```{r, message = FALSE}
subset_merged_df <- convert_location_lat_long(subset_merged_df)
kable(subset_merged_df[1:3, c("location", "lat", "long")])
```

We can see that the values in `location` variable has been split to `lat` and `long`, tidying the dataset.

### Tidy & Manipulate Data II (Variable Creation/Mutation)

In this section, we can extract the `year` and `month` values from `metadata_time`. This is done so that we can focus on the year and month of the observations.

`metadt_to_ind_dt()` function takes metadate_time and creates the `year` and `month` values.

```{r, message = FALSE}
metadt_to_ind_dt <- function(df) {
  df <- df %>%
    mutate(
      year = as.character(year(metadata_time)),
      month = month(metadata_time, label = TRUE)
    )
  return(df)
}
```

```{r, message = FALSE}
subset_merged_df <- metadt_to_ind_dt(subset_merged_df)
kable(subset_merged_df[1:3, c("metadata_time", "year", "month")])
```

### Subset dataset by variables

In this section, we conduct an additional step, by selecting only variables that we want to focus on for the project.
Each variable will be described later.

```{r, message = FALSE}
subset_merged_df <- subset_merged_df %>% select(year, month, name, skip_record_flag, lat, long, payload_fields_temp, payload_fields_hum, payload_fields_battery)
```

### Understand

I will be renaming the columns here so that it can fit within the outputs in the .pdf documentation

```{r, message = FALSE}
subset_merged_df <- rename(subset_merged_df, temp = payload_fields_temp, hum = payload_fields_hum, bat = payload_fields_battery)
```

```{r, message = FALSE}
head(subset_merged_df, n = 3)
```

```{r, message = FALSE}
str(subset_merged_df)
```

Feature           Data Type             Description
--------------    -----------           ------------
year              character             The year of the recorded observation
month             Factor (Ordinal)      The month of the recorded observation
name              character             The name of the location
skip_record_flag  character             A binary categorical variable for skip_record, where "Y" is True and "N" is False
lat               numerical             The latitude of the location
long              numerical             The longitude of the location
temp              numerical             The temperature recorded
hum               numerical             The humidity recorded
bat               numerical             The voltage of the device recorded

#### Data Type Conversions

Some of the variables can be converted to a more specific data types in R, which helps R interpret them better and improve the quality of analysis

Feature           Data Type           Updated Data Type             Reason
--------------    -----------         ------------                  ------
year              character           Factor (Ordinal)              year is an categorical variable that has an order (i.e. 2024 is before 2025)
name              character           Factor                        name is a nominal categorical variable
skip_record_flag  character           Integer                       Relabelled to binary integers to have clarity in analysis and modelling. Similar to One-Hot Encoding, by converting it to integers, this allows models to use numeric (Integer) binary values instead of character labels for the target variable in classification tasks.

Lets take a closer look at `name`

```{r, message = FALSE}
unique(subset_merged_df$name)
```

We can see that there are only a couple of distint `name` values. This means that the `name` is a nominal categorical variable. We will have to convert this from character to factor.

```{r, message = FALSE}
subset_merged_df$name <- factor(subset_merged_df$name)
subset_merged_df$year <- factor(subset_merged_df$year, ordered = TRUE)
```

#### Labelling Factor Variable

We will be relabelling the values, where "Y" is 1 and "N" is 0.

```{r, message = FALSE}
subset_merged_df$skip_record_flag <- ifelse(subset_merged_df$skip_record_flag == "Y", 1L, 0L)
```

#### Final Results after Data Conversion and Labelling

```{r, message = FALSE}
str(subset_merged_df)
```

### Scan I (Addressing Missing, Special and Inconsistent Values)

In this section, we look at common issues found in datasets that have to be cleaned. These include

1) Missing Values: Dataset with missing values are recorded as NA or NaN, and can cause mathematical problems.
2) Special and Inconsistent values: Dataset with such values do not hold the right information, and can cause bias results (e.g. We have an human age value recorded at 1000, but it is impossible for someone to live that long)

#### Missing Values

```{r, message = FALSE}
colSums(is.na(subset_merged_df))
```

There are missing values in `name`, `skip_record_flag`, `lat` and `long`.

The reason why there are missing values in `skip_record_flag` is only one dataset has the `skip_record_flag` variable. During merging, the dataset without this variable would result in the observations having NA for this variable. In this example, we can fill in the missing values, by setting `skip_record_flag` to 0, as we can assume that it may have been ignored since it is not recorded in that dataset.

```{r, message = FALSE}
subset_merged_df$skip_record_flag[is.na(subset_merged_df$skip_record_flag)] <- 0
```

Lets check if the missing `lat` and `long` are the same observations

```{r, message = FALSE}
sum(is.na(subset_merged_df$lat) != is.na(subset_merged_df$long))
```

All observations either have lat and long coordinates, or missing both lat and long coordinates, we can use either variable to analyse or compute the observations with missing values. In this case, we will use lat

```{r, message = FALSE}
cat("No. of Observations ", dim(subset_merged_df)[1], " No. of missing coordinates observations ", sum(is.na(subset_merged_df$lat)))
cat("Percentage of dropped observations ", (sum(is.na(subset_merged_df$lat)) / dim(subset_merged_df)[1]) * 100)
```

This means that we can drop all 968 observations, which would be 8.36% of the observations dropped, slightly above the acceptable range of 5%, but is still generally acceptable. One way to work around this is to add more observations from the original datasets, since our subsetted data is less than 1% of the whole dataset.

```{r, message = FALSE}
subset_merged_df <- subset_merged_df[!(is.na(subset_merged_df$lat)), ]
```

##### After fixing Missing Values

```{r, message = FALSE}
colSums(is.na(subset_merged_df))
```


```{r, message = FALSE}
cat("Dim: ", dim(subset_merged_df))
```

There are no more missing values, and we are now left with 10613 observations.

#### Special and Inconsistent Values

##### Numerical Variables

`get_min_max_values()` function is used to display the min and max values for a numerical variable. We will use this function to see the range of values for all numerical variables.

```{r, message = FALSE}
numerical_variables = c("lat", "long", "temp", "hum", "bat")
get_min_max_values <- function(nums) {
  return(str_c("min: ", min(nums), " max: ", max(nums)))
}
```

- `lat` should be a value between -90 to 90
- `long` should be a value between - 180 to 180
- `hum` should be a value between 0 to 100
- `bat` should be a value above 0, but not too extreme. Voltage should not be too high
- `temp` should be an acceptable range, it cannot be too low (-1000 Celsius is too low) or too hot (1000 Celsius is too high)

```{r, message = FALSE}
for(var in numerical_variables) {
  cat(var, get_min_max_values(subset_merged_df[[var]]), "\n")
}
```

All numerical variables are within acceptable range.

##### Categorical Variables

We will be looking for anomalies in the values (e.g. Incorrect spellings)

```{r, message = FALSE}
nominal_categorical_variables = c("name", "skip_record_flag")
ordinal_categorical_variables = c("year", "month")
```

```{r, message = FALSE}
for(var in nominal_categorical_variables) {
  cat(var, "\n", unique(as.character(subset_merged_df[[var]])), "\n")
}
```

Some observations in `name` only contains the BAT location. We should check out if we can map the BAT locations to an actual location name for better clarity.

```{r, message = FALSE}
bat_loc <- c("BAT-1", "BAT-2", "BAT-3", "BAT-4", "BAT-5")
for(loc in bat_loc) {
  bat_lat_long <- unique(subset_merged_df[subset_merged_df$name == loc, c("lat", "long")])
  result <- unique(subset_merged_df[subset_merged_df$lat == bat_lat_long$lat &
                                subset_merged_df$long == bat_lat_long$long, c("name")])
  cat("BAT locations found with the same coordinates as ", loc ,": ", as.character(result$name), "\n")
}
```

Unfortunately, there are no observations other than itself with the same `lat` and `long`, and cannot determine the actual location name. We will leave the BAT location names as it is for now, but we should replace the names with actual location names to improve readability and consistency throughout the dataset.

```{r, message = FALSE}
for(var in ordinal_categorical_variables) {
  cat(var, "\n", levels(subset_merged_df[[var]]), "\n")
}
```

There are no issues with `year` and `month` variables.

### Scan II (Addressing Outliers)

In this section, we will take a look at outliers that may appear in numeric variables. Identifying outliers are important as they can cause bias in statistical analysis.

We will be looking at both univariate outliers (outliers in a single variable) and multivariate outliers (outliers across multiple variables combined together)

#### Univariate Outliers

Lets take a look at the distribution plot for the variables

```{r, message = FALSE}
par(mfrow = c(2, 3))
hist(subset_merged_df$temp, main = "Distribution Plot for Temperature")
hist(subset_merged_df$hum, main = "Distribution Plot for Humidity")
hist(subset_merged_df$bat, main = "Distribution Plot for Battery")
hist(subset_merged_df$lat, main = "Distribution Plot for Lat")
hist(subset_merged_df$long, main = "Distribution Plot for Long")
```

We will also use Shapiro Test Statistic to check for normality in variables. Do note that we have to subset our dataset to conduct Shaprio Test in R as this works only for datasets with 3 to 5000 observations

```{r, message = FALSE}
for(num_var in numerical_variables) {
  cat(num_var, ": ", shapiro.test(subset_df(subset_merged_df, subset_perc = 0.04)[[num_var]])$p.value, "\n")
}
```

From the histograms, we can see that the distributions are skewed, and all p-values are less than 0.05, therefore the data is not normally distributed. For not normally distributed data, we can use boxplot to determine the outliers.

`display_boxplot_and_get_outlier_indices()` function displays the boxplot and returns the outlier indexes.

```{r, message = FALSE}
display_boxplot_and_get_outlier_indices <- function(var) {
  bp <- boxplot(subset_merged_df[[var]], main = str_c("Boxplot of ", var), ylab = var)
  outlier_indexes <- which(subset_merged_df[[var]] %in% bp$out)
  return(outlier_indexes)
}
```

```{r, message = FALSE}
par(mfrow = c(2, 3))
bp_temp <- display_boxplot_and_get_outlier_indices("temp")
bp_hum <- display_boxplot_and_get_outlier_indices("hum")
bp_bat <- display_boxplot_and_get_outlier_indices("bat")
bp_lat <- display_boxplot_and_get_outlier_indices("lat")
bp_long <- display_boxplot_and_get_outlier_indices("long")
```

While we can drop outliers from the dataset, we can impute the outliers by capping. Capping replaces extreme outliers with a specified percentile value. In `univariate_cap()` function, we cap small value outliers to the 5th percentile and large value outliers to the 95th percentile. This allows us to keep the observations rather than dropping them.

```{r, message = FALSE}
univariate_cap <- function(x){
  quantiles <- quantile(x, c(.05, 0.25, 0.75, .95))
  x[x < quantiles[2] - 1.5 * IQR(x)] <- quantiles[1]
  x[x > quantiles[3] + 1.5 * IQR(x)] <- quantiles[4]
  return(x)
}
```

We will be capping `temp`, `hum` and `bat`. `lat` and `long` should not be capped since they specify the coordinates, and coordinates should not be changed, otherwise, you would be referring to a different location.

```{r, message = FALSE}
subset_merged_df$temp <- univariate_cap(subset_merged_df$temp)
subset_merged_df$hum <- univariate_cap(subset_merged_df$hum)
subset_merged_df$bat <- univariate_cap(subset_merged_df$bat)
```

```{r, message = FALSE}
par(mfrow = c(1, 3))
bp_temp <- display_boxplot_and_get_outlier_indices("temp")
bp_hum <- display_boxplot_and_get_outlier_indices("hum")
bp_bat <- display_boxplot_and_get_outlier_indices("bat")
```

We can see that there are almost no more outliers for `temp`, `hum` and `bat`. It is possible to still detect some outliers in the variables, due to an updated set of quantile values.

#### Multivariate Outliers

We can use `mvn` function to find multivariate outliers. Lets do it for 4 different groups
1) temp and hum
2) temp, hum and bat
3) lat and long
4) All numeric variables

```{r, message = FALSE}
par(mfrow = c(1, 1))
mvn_temp_hum <- mvn(data = subset_merged_df[, c("temp", "hum")], multivariateOutlierMethod = "quan", showOutliers = TRUE)
```

```{r, message = FALSE}
par(mfrow = c(1, 1))
mvn_temp_hum_bat <- mvn(data = subset_merged_df[, c("temp", "hum", "bat")],
                        multivariateOutlierMethod = "quan", showOutliers = TRUE)
```

```{r, message = FALSE}
par(mfrow = c(1, 1))
mvn_lat_long <- mvn(data = subset_merged_df[, c("lat", "long")],
                    multivariateOutlierMethod = "quan", showOutliers = TRUE)
```

```{r, message = FALSE}
par(mfrow = c(1, 1))
mvn_all <- mvn(data = subset_merged_df[, c(numerical_variables)],
                    multivariateOutlierMethod = "quan", showOutliers = TRUE)
```

We can see that there are 815 outliers with `temp` and `hum`, and 1412 outliers with `temp`, `hum` and `bat` combined. Similar to univariate outliers, we should not be removing by `lat` and `long`. In this example, we will be dropping outliers detected for `temp`, `hum` and `bat`.

```{r, message = FALSE}
outlier_indexes <- which(mvn_temp_hum_bat$multivariateOutliers == TRUE)
subset_merged_df <- subset_merged_df[-outlier_indexes, ]
```

```{r, message = FALSE}
cat("Dim: ", dim(subset_merged_df))
```

After removing the outliers detected from `mvn()`, we are left with X observations.

Similar to removing univariate outliers, do note that it is possible to still detect multivariate outliers in the dataset, due to an updated set of quantile values.

```{r, message = FALSE}
par(mfrow = c(1, 1))
mvn_temp_hum_bat <- mvn(data = subset_merged_df[, c("temp", "hum", "bat")],
                        multivariateOutlierMethod = "quan", showOutliers = TRUE)
```

In this case, you can continue remove outliers by repeating the process, until all outliers are removed. Each process would reduce the number of detected outliers, however, affect the total number of observations, as you will be reducing it every time you run an iteration.

For the purpose of keeping as much data as possible, we will only remove the detected multivariate outliers once.

### Transform (Data Transformation)

In this section, we will be applying data transformation to a variable. Lets see how this is done.

```{r, message = FALSE}
par(mfrow = c(2, 2))
hist(subset_merged_df$temp, main = "Distribution Plot for Temperature")
hist(subset_merged_df$hum, main = "Distribution Plot for Humidity")
hist(subset_merged_df$bat, main = "Distribution Plot for Battery")
```

From the histograms, we can see that `hum` is left-skewed. We can conduct data transformation on `hum` to reduce the skewness. A more normally distributed variable can improve the analysis and results of statistical tests (e.g. ANOVA) and models that assume normality (e.g. Linear Regression).

Lets look at some transformation methods
1) Natural Log `log()`
- Applies natural log to the values
2) Log Base 10 `log10()`
- Applies log base 10 to the values
3) Box-Cox Transfomration `BoxCox()`
- Uses an value (lambda) to find the optimal result to normalise the distribution

```{r, message = FALSE}
par(mfrow = c(2, 2))
hist(log(subset_merged_df$hum), main = "Distribution Plot with natural log")
hist(log10(subset_merged_df$hum), main = "Distribution Plot with log base 10")
hist(BoxCox(subset_merged_df$hum, lambda = "auto"), main = "Distribution Plot with BoxCox")
```

We can see that natural log is the best transformation method. Lets transform the values with natural log.

```{r, message = FALSE}
subset_merged_df$hum <- log(subset_merged_df$hum)
```

Lets take a look again after transforming `hum`

```{r, message = FALSE}
par(mfrow = c(1, 1))
hist(subset_merged_df$hum, main = "Distribution Plot for Humidity")
```

### Reflection on Challenges and Ease

- Challenges Faced
  - The most difficult challenge was finding a dataset that met the requirements. They have to be related, and you would also have to imagine how the tidied dataset would look like before determining if the datasets are suitable for the project.
  - Another challenge was figuring out the procedure for preprocessing the dataset, while trying to meet the required components and sections. In this project, we can see that the steps slightly deviated from the original/recommended template, but we still managed to address all preprocessing steps effectively.
- Ease and Strengths
  - The preprocessing steps were generally manageable. I was able to work with the dataset and code well with the guidance provided through lecture and practical lessons.
  - By breaking down each task into smaller subtasks, I was able to organise the structure of the project despite the differences in steps from the template.
- Insights and Learning
  - As much as it was challenging, I learnt that visualising the preprocessing steps while only looking at the datasets is crucial in early decision making
  - I also realised that I was able to figure out the preprocessing methodology easier by reviewing lectures and practicing questions in both lecture and practical sessions. This helped me understand the conceptual knowledge better, and work out the preprocessing code easier (e.g. Scanning for Outliers)

### References
1) Temperature and Humidity for Bats monitoring - Temperature and Humidity for Bats monitoring CSV - Victorian Government Data Directory. (2024). https://discover.data.vic.gov.au/dataset/temperature-and-humidity-for-bats-monitoring/resource/c501ce88-bf15-5e6a-b41a-3e1a145091a3
2) Temperature and Humidity (Rising HF sensors) - Temperature and Humidity (Rising HF sensors) CSV - Victorian Government Data Directory. (2024). https://discover.data.vic.gov.au/dataset/temperature-and-humidity-rising-hf-sensors/resource/14200bdb-fb16-51ba-a312-afac7f89a3d5